---
title: "Supplementary methods and results: parasite-host niche overlap"
author: "Prof. Dr. Bruno Vilela - UFBA | IBio"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    highlight: tango
fontsize: 10pt
geometry: margin=0.75in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{pdflscape}
  - \usepackage{float}
  - \usepackage{xurl}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
  - \setlength{\emergencystretch}{3em}
editor_options:
  chunk_output_type: console
tags:
- Niche overlap
- Broennimann 2012
- Orchid bees
- Parasite-host interactions
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  fig.pos = "H",
  out.width = "95%",
  size = "scriptsize"
)
options(knitr.kable.NA = "", width = 88)
```

# Scope

This supplement is a reproducible tutorial for estimating environmental niche overlap between cleptoparasitic bee species and host bee species. The workflow follows the environmental-space framework of Broennimann et al. (2012): occurrence records are validated, accessible environmental backgrounds are defined, bioclimatic variables are summarized by PCA, species densities are gridded in PCA space, and parasite-host pairs are compared with overlap metrics, randomization tests, and niche dynamics indices.

The document is written to show the analytical decisions needed to reproduce the results. Code chunks display the core analysis logic, while table-formatting and other presentation-only code are hidden. Generated figures and tables are saved locally under `Figures/` and `Results/`, but large intermediate objects are ignored by git so the repository remains lightweight.

# Analysis Setup

Begin by opening `bee_parasites.Rproj`. The project root is then the working directory for all relative paths. Input data are read from `Data/`, reusable functions from `Scripts/`, this document from `Rmarkdown/`, figures are written to `Figures/`, and analytical outputs are written to `Results/`.

The main analytical package is `ecospat`, which implements the Broennimann et al. environmental-space niche comparison functions. Spatial processing uses `terra` and `sf`; figures use `ggplot2`; randomization tests use `ecospat`'s internal randomization routines with reproducible seeds; and `_targets.R` provides a pipeline scaffold for rerunning the complete workflow. The document checks that required packages are installed and stops early if any are missing.

```{r package-requirements-code, eval=FALSE, echo=TRUE}
# Required analytical and reporting packages.
required_packages <- c(
  "ade4", "colorspace", "dplyr", "ecospat", "future", "future.apply",
  "furrr", "geodata", "ggplot2", "ggrepel", "kableExtra", "knitr",
  "patchwork", "purrr", "raster", "readr", "readxl", "rmarkdown",
  "rnaturalearth", "scales", "sessioninfo", "sf", "stringr", "targets",
  "tarchetypes", "terra", "tibble", "tidyr", "viridisLite"
)

missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  stop("Install missing packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r packages}
required_packages <- c(
  "ade4", "colorspace", "dplyr", "ecospat", "future", "future.apply",
  "furrr", "geodata", "ggplot2", "ggrepel", "kableExtra", "knitr",
  "patchwork", "purrr", "raster", "readr", "readxl", "rmarkdown",
  "rnaturalearth", "scales", "sessioninfo", "sf", "stringr", "targets",
  "tarchetypes", "terra", "tibble", "tidyr", "viridisLite"
)

missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install these packages before rendering the supplement.",
    call. = FALSE
  )
}
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r project-setup}
project_dir <- normalizePath(".", mustWork = TRUE)
is_project_root <- function(path) {
  dir.exists(file.path(path, "Data")) &&
    dir.exists(file.path(path, "Rmarkdown")) &&
    length(Sys.glob(file.path(path, "*.Rproj"))) > 0
}
if (!is_project_root(project_dir) && is_project_root(normalizePath("..", mustWork = TRUE))) {
  project_dir <- normalizePath("..", mustWork = TRUE)
}
if (!is_project_root(project_dir)) {
  stop(
    "Render this document from the RStudio Project root. Open the .Rproj file before running analyses.",
    call. = FALSE
  )
}
knitr::opts_knit$set(root.dir = project_dir)

source(file.path(project_dir, "Scripts", "functions", "niche_overlap_functions.R"))
source(file.path(project_dir, "Scripts", "functions", "cache_helpers.R"))
source(file.path(project_dir, "Scripts", "functions", "table_helpers.R"))
source(file.path(project_dir, "Scripts", "analysis", "run_workflow.R"))
create_output_dirs(project_dir)

figure_dir <- file.path(project_dir, "Figures")
tables_dir <- file.path(project_dir, "Results", "tables")
validation_dir <- file.path(project_dir, "Results", "validation")
```

# Data, Settings, and Pair Definition

The occurrence file is `Data/occurrences_bees_parasite_host.csv`. It contains species names, interaction labels (`host` or `parasite`), coordinates, and collection year. The workflow applies a second analytical validation step before extracting environmental values: required columns must be present, coordinates must be numeric and within valid geographic ranges, interaction labels must be recognized, duplicate species-coordinate records are removed, and species must meet the minimum occurrence threshold.

Parasite-host comparisons can be defined in two ways. If `Data/parasite_host_pairs.csv` exists, the workflow uses those curated pairs. If it is absent, all parasite species are crossed with all host species that pass validation. The current run uses the exploratory all-combinations rule, so each comparison should be interpreted as an environmental association test rather than as proof of a confirmed biological interaction.

```{r workflow-settings-code, eval=FALSE, echo=TRUE}
# Workflow settings used for this supplementary render.
workflow_settings <- list(
  occurrence_file = file.path("Data", "occurrences_bees_parasite_host.csv"),
  pair_file = file.path("Data", "parasite_host_pairs.csv"),
  minimum_occurrences = 5,
  buffer_degrees = 2,
  worldclim_resolution = 10,
  grid_resolution = 100,
  run_randomization_tests = TRUE,
  test_repetitions = 100,
  random_seed = 42,
  future_workers = 4,
  future_strategy = "multisession"
)
```

```{r data-loading-code, eval=FALSE, echo=TRUE}
# Data loading and environmental data acquisition.
raw_occurrences <- read_occurrence_file(workflow_settings$occurrence_file)
validation <- validate_occurrence_data(
  raw_occurrences,
  min_occ = workflow_settings$minimum_occurrences,
  validation_dir = file.path("Results", "validation")
)
variables <- load_worldclim_bioclim(
  project_dir = ".",
  resolution = workflow_settings$worldclim_resolution
)
```

```{r settings-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following tables.
settings_report
data_source_table
```

```{r settings, echo=FALSE}
workflow_settings <- list(
  occurrence_file = file.path(project_dir, "Data", "occurrences_bees_parasite_host.csv"),
  pair_file = file.path(project_dir, "Data", "parasite_host_pairs.csv"),
  minimum_occurrences = 5,
  buffer_degrees = 2,
  worldclim_resolution = 10,
  grid_resolution = 100,
  run_randomization_tests = TRUE,
  test_repetitions = 100,
  random_seed = 42,
  future_workers = 4,
  future_strategy = "multisession"
)

settings_report <- tibble::tibble(
  Parameter = c(
    "Minimum cleaned occurrences",
    "Accessible-area buffer",
    "WorldClim resolution",
    "PCA grid resolution",
    "Randomization tests",
    "Test repetitions",
    "Random seed",
    "Parallel workers"
  ),
  Value = c(
    workflow_settings$minimum_occurrences,
    paste(workflow_settings$buffer_degrees, "decimal degrees"),
    paste(workflow_settings$worldclim_resolution, "arc-min"),
    workflow_settings$grid_resolution,
    workflow_settings$run_randomization_tests,
    workflow_settings$test_repetitions,
    workflow_settings$random_seed,
    workflow_settings$future_workers
  )
)

data_source_table <- tibble::tibble(
  Item = c("Occurrence data", "Optional pair definition", "Environmental data"),
  Source = c(
    "Data/occurrences_bees_parasite_host.csv",
    "Data/parasite_host_pairs.csv",
    "WorldClim 2.1 current bioclimatic variables"
  ),
  Use = c(
    "Species identity, interaction category, coordinates, and collection year",
    "Curated parasite-host pairs when available",
    "Environmental predictors used to calibrate PCA niche space"
  )
)

pdf_table(
  settings_report,
  "Analysis settings.",
  font_size = 8,
  widths = c("5.4cm", "4.6cm")
)
pdf_table(
  data_source_table,
  "Input data sources.",
  font_size = 8,
  widths = c("3.4cm", "4.6cm", "6.4cm")
)
```

```{r workflow-execution-code, eval=FALSE, echo=TRUE}
# Complete workflow execution from the project root.
workflow <- run_niche_overlap_workflow(
  project_dir = ".",
  settings = workflow_settings,
  use_cache = TRUE
)

validation <- workflow$validation
pairs <- workflow$pairs
env_space <- workflow$env_space
pca_results <- workflow$pca_results
test_results <- workflow$test_results
figures <- workflow$figures
```

```{r run-workflow}
workflow <- run_niche_overlap_workflow(
  project_dir = project_dir,
  settings = workflow_settings,
  use_cache = TRUE
)

validation <- workflow$validation
pairs <- workflow$pairs
env_space <- workflow$env_space
pca_results <- workflow$pca_results
test_results <- workflow$test_results
figures <- workflow$figures
```

# Analytical Definitions

Niche overlap was calculated in the first two axes of PCA-calibrated environmental space. Let \(P_i\) be the corrected parasite niche density and \(H_i\) be the corrected host niche density in grid cell \(i\). Schoener's D was calculated as

\[
D = 1 - \frac{1}{2}\sum_i |P_i - H_i|.
\]

Warren's I was calculated as

\[
I = 1 - \frac{1}{2}\sum_i \left(\sqrt{P_i} - \sqrt{H_i}\right)^2.
\]

Both indices range from 0 to 1. A value of 0 indicates no environmental overlap, whereas a value of 1 indicates identical environmental occupancy in the calibrated niche space.

For niche dynamics, the host niche was treated as the reference niche and the parasite niche as the focal niche. The categories returned by `ecospat.niche.dyn.index()` were expressed in parasite-host terminology. Shared parasite-host environment corresponds to stability in the original framework and indicates the proportion of parasite density occurring in environmental conditions also occupied by the host. Parasite-exclusive environment corresponds to expansion and indicates the proportion of parasite density outside host-occupied environmental conditions. Host environment unfilled by parasite corresponds to unfilling and indicates the proportion of host density not occupied by the parasite.

Niche equivalency and niche similarity tests are implemented as randomization procedures. Equivalency tests compare observed overlap with null expectations generated by pooling and reallocating occurrences. Similarity tests compare observed relationships with null expectations constrained by the available environmental background. These procedures were executed during rendering with the settings reported below.

# Occurrence Validation

The additional validation stage checked required fields, interaction labels, coordinate format, coordinate ranges, duplicate species-interaction-coordinate records, and minimum cleaned occurrences per species. Records failing validation were excluded from environmental extraction and pairwise comparisons. Complete record-level validation outputs were written to `Results/validation/validation_report.csv`.

```{r occurrence-validation-code, eval=FALSE, echo=TRUE}
# Occurrence validation logic.
required <- c("ID", "Species", "Interaction", "Longitude", "Latitude", "Year_collect")
missing_cols <- setdiff(required, names(raw_occurrences))
if (length(missing_cols) > 0) {
  stop("Occurrence data are missing required columns: ",
       paste(missing_cols, collapse = ", "))
}

prepared <- raw_occurrences |>
  dplyr::mutate(
    Species = stringr::str_squish(as.character(.data$Species)),
    Interaction = tolower(stringr::str_squish(as.character(.data$Interaction))),
    Longitude = parse_numeric_column(.data$Longitude),
    Latitude = parse_numeric_column(.data$Latitude),
    Year_collect = parse_numeric_column(.data$Year_collect)
  )

clean_occurrences <- prepared |>
  dplyr::filter(
    .data$Interaction %in% c("host", "parasite"),
    !is.na(.data$Species),
    !is.na(.data$Longitude),
    !is.na(.data$Latitude),
    dplyr::between(.data$Longitude, -180, 180),
    dplyr::between(.data$Latitude, -90, 90)
  ) |>
  dplyr::distinct(.data$Species, .data$Interaction,
                  round(.data$Longitude, 6), round(.data$Latitude, 6),
                  .keep_all = TRUE)

species_summary <- clean_occurrences |>
  dplyr::count(.data$Interaction, .data$Species, name = "clean_records") |>
  dplyr::mutate(
    analysis_status = dplyr::if_else(
      .data$clean_records >= workflow_settings$minimum_occurrences,
      "eligible",
      "below_minimum_records"
    )
  )
```

```{r validation-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following validation tables.
validation_checks
validation_issue_summary
species_report
```

```{r validation-results, echo=FALSE}
validation_checks <- validation$corrections |>
  dplyr::transmute(
    Check = dplyr::case_when(
      .data$check == "delimiter_and_types" ~ "Delimiter and data types",
      .data$check == "missing_or_non_numeric_coordinates" ~ "Missing or non-numeric coordinates",
      .data$check == "coordinates_outside_valid_range" ~ "Coordinates outside valid range",
      .data$check == "invalid_interaction_labels" ~ "Invalid interaction labels",
      .data$check == "duplicate_species_interaction_coordinate" ~ "Duplicate species-coordinate records",
      .data$check == "minimum_occurrence_threshold" ~ "Minimum occurrence threshold",
      TRUE ~ .data$check
    ),
    Affected = .data$records_affected,
    Action = wrap_table_text(.data$correction_applied, 70)
  )

species_report <- validation$species_summary |>
  dplyr::transmute(
    Interaction = .data$Interaction,
    Species = species_label(.data$Species),
    Raw = .data$raw_records,
    Clean = .data$clean_records,
    Duplicates = .data$duplicates_removed,
    Removed = .data$invalid_or_missing_removed,
    Status = dplyr::if_else(
      .data$analysis_status == "below_minimum_records",
      "below minimum",
      .data$analysis_status
    )
  )

validation_issue_summary <- validation$validation_report |>
  dplyr::count(.data$validation_issue, name = "Records") |>
  dplyr::mutate(
    Issue = stringr::str_to_sentence(stringr::str_replace_all(.data$validation_issue, "_", " "))
  ) |>
  dplyr::select(Issue, Records)
if (nrow(validation_issue_summary) == 0) {
  validation_issue_summary <- tibble::tibble(Issue = "No record-level validation issues detected", Records = 0L)
}

pdf_table(
  validation_checks,
  "Validation checks and corrective actions.",
  font_size = 7.2,
  widths = c("4.0cm", "1.4cm", "9.2cm")
)
pdf_table(
  validation_issue_summary,
  "Record-level validation issue summary.",
  font_size = 7.5,
  widths = c("9.0cm", "2.0cm")
)
pdf_table(
  species_report,
  "Species-level occurrence counts after validation.",
  font_size = 6.7,
  longtable = TRUE,
  widths = c("1.6cm", "4.0cm", "0.9cm", "0.9cm", "1.4cm", "1.1cm", "2.8cm")
)
```

```{r fig-occurrences, fig.cap="Validated host and parasite occurrence records retained after coordinate screening and duplicate filtering. Colours and symbols distinguish interaction category; facets identify species.", fig.width=10, fig.height=8}
figures$fig_occurrences
```

# Pair Definition and Environmental Data

```{r pair-generation-code, eval=FALSE, echo=TRUE}
# Parasite-host pair generation.
parasite_counts <- species_summary |>
  dplyr::filter(.data$Interaction == "parasite") |>
  dplyr::select(parasite = Species, n_parasite = clean_records)
host_counts <- species_summary |>
  dplyr::filter(.data$Interaction == "host") |>
  dplyr::select(host = Species, n_host = clean_records)

if (file.exists(workflow_settings$pair_file)) {
  raw_pairs <- read_occurrence_file(workflow_settings$pair_file)
  candidate_pairs <- raw_pairs |>
    dplyr::transmute(
      parasite = gsub(" ", "_", stringr::str_squish(.data$parasite)),
      host = gsub(" ", "_", stringr::str_squish(.data$host)),
      pairing_rule = "user_supplied_pair_file"
    ) |>
    dplyr::distinct()
} else {
  parasite_species <- species_summary |>
    dplyr::filter(.data$Interaction == "parasite") |>
    dplyr::pull(.data$Species)
  host_species <- species_summary |>
    dplyr::filter(.data$Interaction == "host") |>
    dplyr::pull(.data$Species)
  candidate_pairs <- tidyr::expand_grid(
    parasite = parasite_species,
    host = host_species
  ) |>
    dplyr::mutate(pairing_rule = "all_candidate_parasite_host_combinations")
}

valid_pairs <- candidate_pairs |>
  dplyr::left_join(parasite_counts, by = "parasite") |>
  dplyr::left_join(host_counts, by = "host") |>
  dplyr::filter(
    .data$n_parasite >= workflow_settings$minimum_occurrences,
    .data$n_host >= workflow_settings$minimum_occurrences
  )
```

```{r environmental-extraction-code, eval=FALSE, echo=TRUE}
# Species-specific accessible areas and environmental extraction.
species_to_analyze <- unique(c(valid_pairs$parasite, valid_pairs$host))
env_space <- prepare_environment_space(
  clean_occ = validation$clean,
  variables = variables,
  species_to_analyze = species_to_analyze,
  buffer_size = workflow_settings$buffer_degrees,
  min_occ = workflow_settings$minimum_occurrences
)

pairs_environment <- filter_pairs_by_environment(
  valid_pairs,
  env_space$species_summary
)
```

```{r pair-environment-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following pair and environment tables.
pair_summary
environment_report
```

```{r pair-environment-results, echo=FALSE}
pairing_rule <- unique(pairs$all$pairing_rule)
pairing_label <- if (length(pairing_rule) == 1 &&
                     pairing_rule == "user_supplied_pair_file") {
  "Curated pair file"
} else {
  "All possible parasite-host species combinations represented in the occurrence dataset"
}

pair_summary <- tibble::tibble(
  Item = c(
    "Candidate pairs",
    "Valid analyzed pairs",
    "Skipped pairs",
    "Pairing source"
  ),
  Value = c(
    nrow(pairs$all),
    nrow(pairs$valid),
    nrow(pairs$skipped),
    wrap_table_text(pairing_label, 58)
  )
)

environment_report <- env_space$species_summary |>
  dplyr::transmute(
    Species = .data$Species_display,
    Occurrences = .data$occurrence_records_for_environment,
    `Background cells` = .data$background_cells,
    Status = .data$environmental_status
  )

pdf_table(
  pair_summary,
  "Parasite-host pair definition.",
  font_size = 7.8,
  widths = c("5.2cm", "7.4cm")
)
pdf_table(
  environment_report,
  "Environmental extraction and accessible-area background summary.",
  font_size = 7.2,
  longtable = TRUE,
  widths = c("4.4cm", "2.2cm", "2.5cm", "4.6cm")
)
```

Environmental availability was approximated separately for each species. Minimum convex polygons were constructed from validated occurrences, expanded by a two-degree buffer, and clipped to terrestrial environments. These accessible areas are operational approximations of available environments rather than definitive dispersal-accessible regions. This choice preserves compatibility with the original workflow while making the assumption explicit.

```{r fig-backgrounds, fig.cap="Species-specific accessible-area backgrounds used for environmental extraction. Polygons were defined from minimum convex polygons around validated occurrences, expanded by a two-degree buffer, and clipped to land.", fig.width=10, fig.height=8}
figures$fig_backgrounds
```

# Environmental PCA

The next step is to build the environmental space used for niche comparison. The PCA is calibrated on background environments sampled from the species-specific accessible areas, and occurrence records are projected into that same space. This keeps the comparison in environmental space rather than geographic space.

Only the first two PCA axes are used for `ecospat.grid.clim.dyn()`. In this run, those axes capture the main climatic gradients available to the analyzed hosts and parasites.

```{r pca-calibration-code, eval=FALSE, echo=TRUE}
# PCA calibration in environmental space.
all_spec_env <- dplyr::bind_rows(env_space$spec_env)
all_back_env <- dplyr::bind_rows(env_space$back_env)
data_env <- dplyr::bind_rows(all_spec_env, all_back_env)

# Occurrences are projected into the PCA; background environments calibrate it.
row_weights <- c(rep(0, nrow(all_spec_env)), rep(1, nrow(all_back_env)))
pca <- ade4::dudi.pca(
  data_env,
  row.w = row_weights,
  center = TRUE,
  scale = TRUE,
  scannf = FALSE,
  nf = 2
)

pca_scores_occurrences <- pca$li[seq_len(nrow(all_spec_env)), 1:2]
pca_scores_background <- pca$li[-seq_len(nrow(all_spec_env)), 1:2]
pca_loadings <- pca$co[, 1:2]
```

```{r density-grid-code, eval=FALSE, echo=TRUE}
# Broennimann et al. (2012) density grids in PCA environmental space.
grids <- lapply(names(pca_results$scores_spec), function(sp_name) {
  ecospat::ecospat.grid.clim.dyn(
    glob = pca_results$total_scores_back,
    glob1 = pca_results$scores_back[[sp_name]],
    sp = pca_results$scores_spec[[sp_name]],
    R = workflow_settings$grid_resolution
  )
})
names(grids) <- names(pca_results$scores_spec)
```

```{r pca-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following PCA tables.
pca_eigen_report
pca_extreme_report
```

```{r pca-results, echo=FALSE}
bioclim_lookup <- tibble::tibble(
  variable = paste0("bio", 1:19),
  definition = c(
    "Annual mean temperature",
    "Mean diurnal range",
    "Isothermality",
    "Temperature seasonality",
    "Maximum temperature of warmest month",
    "Minimum temperature of coldest month",
    "Temperature annual range",
    "Mean temperature of wettest quarter",
    "Mean temperature of driest quarter",
    "Mean temperature of warmest quarter",
    "Mean temperature of coldest quarter",
    "Annual precipitation",
    "Precipitation of wettest month",
    "Precipitation of driest month",
    "Precipitation seasonality",
    "Precipitation of wettest quarter",
    "Precipitation of driest quarter",
    "Precipitation of warmest quarter",
    "Precipitation of coldest quarter"
  ),
  domain = c(rep("temperature", 11), rep("precipitation", 8))
)

pca_eigen_report <- pca_results$eigenvalues |>
  dplyr::slice_head(n = 5) |>
  dplyr::transmute(
    Axis = .data$axis,
    Eigenvalue = .data$eigenvalue,
    `Variance (%)` = .data$variance_percent
  )

pca_loading_report <- pca_results$loadings |>
  dplyr::mutate(
    absolute_contribution = sqrt(.data$axis1^2 + .data$axis2^2)
  ) |>
  dplyr::left_join(bioclim_lookup, by = "variable") |>
  dplyr::arrange(dplyr::desc(.data$absolute_contribution)) |>
  dplyr::transmute(
    Variable = .data$variable,
    Definition = wrap_table_text(.data$definition, 42),
    PC1 = .data$axis1,
    PC2 = .data$axis2,
    `Absolute contribution` = .data$absolute_contribution
  )

axis_extremes <- function(loadings, lookup, axis, label) {
  positive <- loadings |>
    dplyr::arrange(dplyr::desc(.data[[axis]])) |>
    dplyr::slice_head(n = 3) |>
    dplyr::mutate(Direction = "Positive")
  negative <- loadings |>
    dplyr::arrange(.data[[axis]]) |>
    dplyr::slice_head(n = 3) |>
    dplyr::mutate(Direction = "Negative")
  dplyr::bind_rows(positive, negative) |>
    dplyr::left_join(lookup, by = "variable") |>
    dplyr::transmute(
      Axis = label,
      Direction = .data$Direction,
      Variable = .data$variable,
      Definition = wrap_table_text(.data$definition, 40),
      Loading = .data[[axis]]
    )
}

pca_extreme_report <- dplyr::bind_rows(
  axis_extremes(pca_results$loadings, bioclim_lookup, "axis1", "PC1"),
  axis_extremes(pca_results$loadings, bioclim_lookup, "axis2", "PC2")
)

pc1_variance <- pca_results$eigenvalues$variance_percent[[1]]
pc2_variance <- pca_results$eigenvalues$variance_percent[[2]]

pdf_table(
  pca_eigen_report,
  "First five PCA eigenvalues and variance explained.",
  digits = 2,
  font_size = 8
)
pdf_table(
  pca_extreme_report,
  "Strongest positive and negative PCA loadings.",
  digits = 3,
  font_size = 7,
  widths = c("1.1cm", "1.8cm", "1.4cm", "6.8cm", "1.6cm")
)
```

PC1 explained `r sprintf("%.1f", pc1_variance)`% of the environmental variation and separated warm, moist, weakly seasonal environments from more thermally variable and comparatively cooler or drier environments. Positive PC1 loadings were strongest for minimum temperature of the coldest month, mean temperature of the driest and coldest quarters, annual mean temperature, and annual precipitation. Negative PC1 loadings were strongest for temperature annual range, mean diurnal range, and temperature seasonality.

PC2 explained `r sprintf("%.1f", pc2_variance)`% of the environmental variation and separated hot, precipitation-seasonal environments with dry dry-season conditions from the rest of the background space. Positive PC2 loadings were strongest for maximum temperature of the warmest month, mean temperatures of warm or wet quarters, and precipitation seasonality. Negative PC2 loadings were strongest for precipitation of the driest quarter, driest month, warmest quarter, and annual precipitation. Therefore, high PC2 values represent hot and seasonally dry conditions, whereas low PC2 values represent environments with greater dry-season moisture.

```{r fig-pca-loadings, fig.cap="Loadings of WorldClim bioclimatic variables on the first two PCA axes used to define environmental niche space.", fig.width=7, fig.height=6}
figures$fig_loadings
```

# Niche Overlap Metrics

After the density grids are created, each parasite-host pair is compared in PCA space. Schoener's D is the primary overlap metric reported here because it is widely used in Broennimann-style niche comparisons. Warren's I is also calculated as a complementary similarity index. For both metrics, 0 indicates no overlap and 1 indicates identical environmental occupancy in the calibrated environmental space.

```{r overlap-calculation-code, eval=FALSE, echo=TRUE}
# Pairwise niche overlap in environmental space.
pair_metrics <- purrr::map_dfr(seq_len(nrow(pairs$valid)), function(i) {
  pair <- pairs$valid[i, ]
  z_parasite <- grids[[pair$parasite]]
  z_host <- grids[[pair$host]]
  overlap <- ecospat::ecospat.niche.overlap(z_parasite, z_host, cor = TRUE)

  tibble::tibble(
    pair_id = pair$pair_id,
    parasite = pair$parasite,
    host = pair$host,
    schoener_d = unname(overlap[["D"]]),
    warren_i = unname(overlap[["I"]])
  )
})
```

```{r overlap-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following overlap tables.
overlap_report
overlap_summary
```

```{r overlap-results, echo=FALSE}
overlap_report <- test_results$metrics |>
  dplyr::arrange(dplyr::desc(.data$schoener_d)) |>
  dplyr::transmute(
    Parasite = wrap_table_text(.data$parasite_label, 20),
    Host = wrap_table_text(.data$host_label, 20),
    `n parasite` = .data$n_parasite,
    `n host` = .data$n_host,
    `Schoener D` = .data$schoener_d,
    `Warren I` = .data$warren_i
  )

highest_overlap <- test_results$metrics |>
  dplyr::slice_max(.data$schoener_d, n = 1, with_ties = FALSE)
lowest_overlap <- test_results$metrics |>
  dplyr::slice_min(.data$schoener_d, n = 1, with_ties = FALSE)
overlap_summary <- tibble::tibble(
  Metric = c("Mean Schoener D", "Range Schoener D", "Mean Warren I", "Range Warren I"),
  Value = c(
    sprintf("%.3f", mean(test_results$metrics$schoener_d, na.rm = TRUE)),
    sprintf(
      "%.3f-%.3f",
      min(test_results$metrics$schoener_d, na.rm = TRUE),
      max(test_results$metrics$schoener_d, na.rm = TRUE)
    ),
    sprintf("%.3f", mean(test_results$metrics$warren_i, na.rm = TRUE)),
    sprintf(
      "%.3f-%.3f",
      min(test_results$metrics$warren_i, na.rm = TRUE),
      max(test_results$metrics$warren_i, na.rm = TRUE)
    )
  )
)

pdf_table(
  overlap_report,
  "Pairwise niche overlap metrics.",
  digits = 3,
  font_size = 7.2,
  longtable = TRUE,
  landscape = TRUE,
  widths = c("4.3cm", "4.5cm", "1.7cm", "1.5cm", "2.0cm", "1.9cm")
)
pdf_table(
  overlap_summary,
  "Summary of niche overlap values.",
  font_size = 8,
  widths = c("4.5cm", "3.0cm")
)
```

The highest Schoener's D was observed for *`r highest_overlap$parasite_label`* versus *`r highest_overlap$host_label`* (`r paste0("D = ", sprintf("%.3f", highest_overlap$schoener_d))`). The lowest Schoener's D was observed for *`r lowest_overlap$parasite_label`* versus *`r lowest_overlap$host_label`* (`r paste0("D = ", sprintf("%.3f", lowest_overlap$schoener_d))`). This range shows that some parasite-host comparisons occupy similar environments, whereas others are nearly separated in the calibrated climatic space.

```{r fig-overlap, fig.cap="Pairwise Schoener's D values for all valid parasite-host comparisons. Darker colours indicate higher environmental niche overlap.", fig.width=8.5, fig.height=5.5}
figures$fig_overlap
```

Randomization tests add an inferential comparison to the observed overlap values. The equivalency test pools parasite and host occurrences, reallocates them randomly, and asks whether the observed overlap is consistent with interchangeable environmental occupancies. The similarity test keeps the available environmental background explicit and asks whether one species is more similar to the other than expected given the available background. Similarity is directional, so the workflow runs parasite-to-host and host-to-parasite tests.

Randomization p-values are calculated from finite null replicates. In rare cases, a randomized replicate can produce an empty or degenerate density grid and return `NA`; those replicates are omitted from the p-value denominator, and the number of valid equivalency repetitions is reported in the p-value table.

```{r randomization-configuration-code, eval=FALSE, echo=TRUE}
# Randomization configuration used in this render.
workflow_settings$run_randomization_tests <- TRUE
workflow_settings$test_repetitions <- 100
workflow_settings$random_seed <- 42
workflow_settings$future_workers <- 4

if (isTRUE(workflow_settings$run_randomization_tests) &&
    workflow_settings$test_repetitions > 0) {
  ecospat_cores <- workflow_settings$future_workers

  equivalency <- ecospat::ecospat.niche.equivalency.test(
    z_parasite,
    z_host,
    rep = workflow_settings$test_repetitions,
    ncores = ecospat_cores
  )
  similarity_parasite_to_host <- ecospat::ecospat.niche.similarity.test(
    z_parasite,
    z_host,
    rep = workflow_settings$test_repetitions,
    ncores = ecospat_cores
  )
  similarity_host_to_parasite <- ecospat::ecospat.niche.similarity.test(
    z_host,
    z_parasite,
    rep = workflow_settings$test_repetitions,
    ncores = ecospat_cores
  )
}
```

```{r randomization-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following randomization tables.
randomization_report
randomization_p_report
randomization_summary
```

```{r randomization-status, echo=FALSE}
randomization_report <- tibble::tibble(
  Test = c(
    "Niche equivalency",
    "Niche similarity: parasite to host",
    "Niche similarity: host to parasite"
  ),
  Status = ifelse(workflow_settings$run_randomization_tests, "Run", "Not run"),
  Repetitions = workflow_settings$test_repetitions,
  Output = c(
    "Results/tables/niche_equivalency_null_models.csv",
    "Results/tables/niche_similarity_null_models.csv",
    "Results/tables/niche_similarity_null_models.csv"
  )
)

randomization_p_report <- test_results$metrics |>
  dplyr::arrange(.data$equivalency_p_d, .data$similarity_p_d_parasite_to_host) |>
  dplyr::transmute(
    Parasite = wrap_table_text(.data$parasite_label, 20),
    Host = wrap_table_text(.data$host_label, 20),
    `Observed D` = .data$schoener_d,
    `Eq. D p` = .data$equivalency_p_d,
    `Eq. valid reps` = .data$equivalency_valid_repetitions,
    `Sim. D P->H p` = .data$similarity_p_d_parasite_to_host,
    `Sim. D H->P p` = .data$similarity_p_d_host_to_parasite
  )

randomization_summary <- tibble::tibble(
  Test = c(
    "Equivalency, Schoener D",
    "Equivalency, Warren I",
    "Similarity D, parasite to host",
    "Similarity D, host to parasite",
    "Similarity I, parasite to host",
    "Similarity I, host to parasite"
  ),
  `Pairs with p < 0.05` = c(
    sum(test_results$metrics$equivalency_p_d < 0.05, na.rm = TRUE),
    sum(test_results$metrics$equivalency_p_i < 0.05, na.rm = TRUE),
    sum(test_results$metrics$similarity_p_d_parasite_to_host < 0.05, na.rm = TRUE),
    sum(test_results$metrics$similarity_p_d_host_to_parasite < 0.05, na.rm = TRUE),
    sum(test_results$metrics$similarity_p_i_parasite_to_host < 0.05, na.rm = TRUE),
    sum(test_results$metrics$similarity_p_i_host_to_parasite < 0.05, na.rm = TRUE)
  ),
  `Pairs tested` = nrow(test_results$metrics),
  Repetitions = workflow_settings$test_repetitions,
  `Minimum valid reps` = c(
    min(test_results$metrics$equivalency_valid_repetitions, na.rm = TRUE),
    min(test_results$metrics$equivalency_valid_repetitions, na.rm = TRUE),
    min(test_results$metrics$similarity_valid_repetitions_parasite_to_host, na.rm = TRUE),
    min(test_results$metrics$similarity_valid_repetitions_host_to_parasite, na.rm = TRUE),
    min(test_results$metrics$similarity_valid_repetitions_parasite_to_host, na.rm = TRUE),
    min(test_results$metrics$similarity_valid_repetitions_host_to_parasite, na.rm = TRUE)
  )
)

pdf_table(
  randomization_report,
  "Randomization test status.",
  font_size = 7.2,
  widths = c("4.3cm", "2.0cm", "1.6cm", "6.2cm")
)
pdf_table(
  randomization_summary,
  "Summary of randomization-test results.",
  font_size = 7.2,
  widths = c("4.8cm", "2.4cm", "1.7cm", "1.7cm", "2.3cm")
)
pdf_table(
  randomization_p_report,
  "Pairwise Schoener's D randomization-test p-values.",
  digits = 3,
  font_size = 7.0,
  longtable = TRUE,
  landscape = TRUE,
  widths = c("4.2cm", "4.5cm", "1.8cm", "1.6cm", "2.0cm", "2.2cm", "2.2cm")
)
```

# Parasite-Host Niche Dynamics

Niche dynamics metrics partition the parasite and host density grids into interpretable components. In invasion studies, `ecospat` labels these components as stability, expansion, and unfilling. Here the mathematical definitions are retained, but the biological wording is changed for parasite-host comparisons.

Use the host as the reference niche and the parasite as the focal niche. Stability is therefore reported as shared parasite-host environment: the proportion of parasite environmental density that falls inside host-occupied environmental conditions. Expansion is reported as parasite-exclusive environment: the proportion of parasite density outside the host-occupied environment. Unfilling is reported as host environment unfilled by parasite: the proportion of host environmental density not occupied by the parasite.

```{r niche-dynamics-code, eval=FALSE, echo=TRUE}
# Niche dynamics with host as the reference and parasite as the focal niche.
pair_dynamic_metrics <- purrr::map_dfr(seq_len(nrow(pairs$valid)), function(i) {
  pair <- pairs$valid[i, ]
  z_parasite <- grids[[pair$parasite]]
  z_host <- grids[[pair$host]]

  # ecospat labels are retained mathematically and relabeled biologically below.
  dyn_host_reference <- ecospat::ecospat.niche.dyn.index(
    z_host,
    z_parasite
  )$dynamic.index.w

  tibble::tibble(
    pair_id = pair$pair_id,
    parasite_host_shared_stability = unname(dyn_host_reference[["stability"]]),
    parasite_exclusive_environment_use = unname(dyn_host_reference[["expansion"]]),
    host_environment_unfilled_by_parasite = unname(dyn_host_reference[["unfilling"]])
  )
})
```

```{r dynamics-reported-objects-code, eval=FALSE, echo=TRUE}
# Objects summarized in the following niche-dynamics tables.
dynamic_report
```

```{r dynamics-results, echo=FALSE}
dynamic_report <- test_results$metrics |>
  dplyr::arrange(dplyr::desc(.data$schoener_d)) |>
  dplyr::transmute(
    Parasite = wrap_table_text(.data$parasite_label, 20),
    Host = wrap_table_text(.data$host_label, 20),
    `Shared environment (stability)` = .data$parasite_host_shared_stability,
    `Parasite-exclusive environment (expansion)` = .data$parasite_exclusive_environment_use,
    `Host unfilled environment (unfilling)` = .data$host_environment_unfilled_by_parasite
  )

highest_shared <- test_results$metrics |>
  dplyr::slice_max(.data$parasite_host_shared_stability, n = 1, with_ties = FALSE)
highest_exclusive <- test_results$metrics |>
  dplyr::slice_max(.data$parasite_exclusive_environment_use, n = 1, with_ties = FALSE)
highest_unfilled <- test_results$metrics |>
  dplyr::slice_max(.data$host_environment_unfilled_by_parasite, n = 1, with_ties = FALSE)

pdf_table(
  dynamic_report,
  "Parasite-host niche dynamics metrics.",
  digits = 3,
  font_size = 7.0,
  longtable = TRUE,
  landscape = TRUE,
  widths = c("4.2cm", "4.5cm", "3.0cm", "3.4cm", "3.4cm")
)
```

The highest shared parasite-host environment was observed for *`r highest_shared$parasite_label`* versus *`r highest_shared$host_label`* (`r sprintf("%.3f", highest_shared$parasite_host_shared_stability)`). The highest parasite-exclusive environment was observed for *`r highest_exclusive$parasite_label`* versus *`r highest_exclusive$host_label`* (`r sprintf("%.3f", highest_exclusive$parasite_exclusive_environment_use)`). The highest host environment unfilled by parasite was observed for *`r highest_unfilled$parasite_label`* versus *`r highest_unfilled$host_label`* (`r sprintf("%.3f", highest_unfilled$host_environment_unfilled_by_parasite)`). These summaries identify pairs with the strongest environmental sharing, the greatest parasite environmental exclusivity, and the largest portion of host environmental space not occupied by the parasite.

```{r fig-dynamics-bars, fig.cap="Parasite-host niche dynamics metrics for each valid pair, sorted by Schoener's D. Values are shown on the original 0-1 scale and are not normalized across metrics.", fig.width=8.8, fig.height=8.6}
figures$fig_dynamic_metrics
```

```{r fig-niche-space, fig.cap="Niche dynamics in PCA environmental space for the pair with the highest Schoener's D. Filled cells show only dynamic occupancy categories; unclassified available background cells are left blank. Host and parasite outlines use solid lines for the full occupied niche support and dashed lines for the 50% density contour.", fig.width=7.5, fig.height=6.5}
figures$fig_top_dynamics
```

# Reproducibility and Output Files

The workflow is reproducible from the RStudio Project root. Deterministic steps use fixed settings, randomization procedures use fixed seeds, and intermediate products are cached by analytical stage. To rerun from scratch, remove `Results/cache/` and rerender this document or run the `targets` pipeline.

```{r caching-configuration-code, eval=FALSE, echo=TRUE}
# Analytical stages cached by explicit input fingerprints.
validation_cache <- cache_eval(
  cache_dir = file.path("Results", "cache"),
  name = "occurrence_cleaning",
  key = list(
    occurrence = input_fingerprint(workflow_settings$occurrence_file),
    minimum_occurrences = workflow_settings$minimum_occurrences
  ),
  code = {
    raw_occurrences <- read_occurrence_file(workflow_settings$occurrence_file)
    validate_occurrence_data(raw_occurrences,
                             min_occ = workflow_settings$minimum_occurrences,
                             validation_dir = file.path("Results", "validation"))
  },
  use_cache = TRUE
)

cached_stages <- c(
  "occurrence_cleaning",
  "pair_definition",
  "environmental_extraction_accessible_areas",
  "environmental_pca",
  "niche_density_grids",
  "overlap_dynamics_randomization",
  "figure_generation"
)
```

```{r figure-generation-code, eval=FALSE, echo=TRUE}
# Publication figures generated by the workflow.
fig_occurrences <- plot_occurrence_map(validation$clean, env_space$world)
fig_backgrounds <- plot_background_map(
  env_space$background_polygons,
  validation$clean,
  env_space$world
)
fig_overlap <- plot_overlap_heatmap(test_results$metrics)
fig_dynamic_metrics <- plot_dynamic_metrics_dotplot(test_results$metrics)
fig_loadings <- plot_pca_loadings(pca_results$loadings, pca_results$eigenvalues)
top_pair <- test_results$metrics |>
  dplyr::slice_max(.data$schoener_d, n = 1, with_ties = FALSE)
fig_top_dynamics <- plot_niche_dynamics_pair(top_pair, grids)

save_ggplot_dual(fig_occurrences, file.path("Figures", "Figure1_cleaned_occurrences"),
                 width = 10, height = 8)
save_ggplot_dual(fig_backgrounds, file.path("Figures", "Figure2_accessible_area_backgrounds"),
                 width = 10, height = 8)
save_ggplot_dual(fig_overlap, file.path("Figures", "Figure3_schoener_d_heatmap"),
                 width = 8.5, height = 5.5)
save_ggplot_dual(fig_dynamic_metrics, file.path("Figures", "Figure4_niche_dynamics_metrics"),
                 width = 8.6, height = 9.0)
save_ggplot_dual(fig_top_dynamics, file.path("Figures", "Figure5_niche_dynamics_top_pair"),
                 width = 7.5, height = 6.5)
save_ggplot_dual(fig_loadings, file.path("Figures", "Figure6_pca_loadings"),
                 width = 7, height = 6)
save_niche_dynamic_plots(test_results$metrics, grids, "Figures")
```

```{r targets-configuration-code, eval=FALSE, echo=TRUE}
# targets scaffold for full workflow execution.
library(targets)
library(tarchetypes)

tar_option_set(
  packages = c("ade4", "dplyr", "ecospat", "future", "future.apply",
               "furrr", "geodata", "ggplot2", "sf", "terra", "tidyr")
)

list(
  tar_target(workflow_settings, list(
    minimum_occurrences = 5,
    buffer_degrees = 2,
    worldclim_resolution = 10,
    grid_resolution = 100,
    run_randomization_tests = TRUE,
    test_repetitions = 100,
    random_seed = 42,
    future_workers = 4,
    future_strategy = "multisession"
  )),
  tar_target(
    workflow_results,
    run_niche_overlap_workflow(".", workflow_settings, use_cache = TRUE)
  ),
  tar_render(
    supplementary_pdf,
    "Rmarkdown/parasite_host_bees.Rmd",
    output_file = "parasite_host_niche_overlap_supplement.pdf",
    output_dir = "Results"
  )
)
```

```{r reproducibility-reported-objects-code, eval=FALSE, echo=TRUE}
# Object summarized in the following reproducibility table.
reproducibility_summary
```

```{r reproducibility-output, echo=FALSE}
parasite_species <- validation$species_summary |>
  dplyr::filter(.data$Interaction == "parasite", .data$analysis_status == "eligible") |>
  dplyr::pull(.data$Species) |>
  species_label()
host_species <- validation$species_summary |>
  dplyr::filter(.data$Interaction == "host", .data$analysis_status == "eligible") |>
  dplyr::pull(.data$Species) |>
  species_label()

reproducibility_summary <- tibble::tibble(
  Item = c(
    "Parasite species analyzed",
    "Host species analyzed",
    "Valid pairwise comparisons",
    "Environmental variables",
    "Randomization status",
    "Cache directory"
  ),
  Value = c(
    length(parasite_species),
    length(host_species),
    nrow(pairs$valid),
    "WorldClim bio1-bio19",
    ifelse(workflow_settings$run_randomization_tests, "randomization tests run", "randomization tests not run"),
    "Results/cache"
  )
)

pdf_table(
  reproducibility_summary,
  "Reproducibility summary.",
  font_size = 7.8,
  widths = c("5.2cm", "7.2cm")
)
```

The main generated outputs are `Results/validation/` for validation logs, `Results/tables/` for analytical tables and null-model outputs, `Figures/` for publication figures, `Figures/Niche_Plots/` for pair-specific niche-space plots, and `Results/session_info.txt` for the full computational environment.

# Computational Environment

```{r computational-environment-reported-object-code, eval=FALSE, echo=TRUE}
# Object summarized in the following computational-environment table.
session_report
```

```{r computational-environment, echo=FALSE}
session_info_file <- file.path(project_dir, "Results", "session_info.txt")
capture.output(sessioninfo::session_info(), file = session_info_file)

pkg_version <- function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    as.character(utils::packageVersion(pkg))
  } else {
    "not installed"
  }
}

session_report <- tibble::tibble(
  Component = c(
    "R", "OS", "ecospat", "terra", "raster", "ggplot2",
    "future", "targets"
  ),
  Version = c(
    R.version.string,
    paste(Sys.info()[["sysname"]], Sys.info()[["release"]]),
    pkg_version("ecospat"),
    pkg_version("terra"),
    pkg_version("raster"),
    pkg_version("ggplot2"),
    pkg_version("future"),
    pkg_version("targets")
  )
)

pdf_table(
  session_report,
  "Computational environment summary.",
  font_size = 7.5,
  widths = c("4.0cm", "9.8cm")
)
```

Full session information was exported to `Results/session_info.txt`.

# Code Availability

All scripts, functions, workflow files, and rendering materials are available from the project repository: <https://github.com/BrunoVilela/parasite_host_bees>.

The repository is organized as follows: `Data/` contains occurrence inputs and optional curated pair definitions; `Scripts/functions/` contains reusable analysis, plotting, table, and cache helpers; `Scripts/analysis/` contains the executable workflow wrapper; `Rmarkdown/` contains this supplementary document; `Figures/` contains generated figure outputs; and `Results/` contains generated tables, validation summaries, and reproducibility outputs. The project should be opened and executed through the included `.Rproj` file so that all relative paths resolve from the project root.

# References

Broennimann, O., Fitzpatrick, M. C., Pearman, P. B., Petitpierre, B., Pellissier, L., Yoccoz, N. G., Thuiller, W., Fortin, M.-J., Randin, C., Zimmermann, N. E., Graham, C. H., & Guisan, A. (2012). Measuring ecological niche overlap from occurrence and spatial environmental data. *Global Ecology and Biogeography*, 21(4), 481-497.

Di Cola, V., Broennimann, O., Petitpierre, B., Breiner, F. T., D'Amen, M., Randin, C., Engler, R., Pottier, J., Pio, D., Dubuis, A., Pellissier, L., Mateo, R. G., Hordijk, W., Salamin, N., & Guisan, A. (2017). ecospat: an R package to support spatial analyses and modeling of species niches and distributions. *Ecography*, 40(6), 774-787.

Silva, D. P., Varela, S., Nemesio, A., & De Marco, P. Jr. (2015). Adding biotic interactions into paleodistribution models: a host-cleptoparasite complex of Neotropical orchid bees. *PLOS ONE*, 10(6), e0129890. https://doi.org/10.1371/journal.pone.0129890
